Tutorial: Quantifying Spatially Resolved Hydration Thermodynamics Using Grid Inhomogeneous Solvation Theory [Article v\versionnumber]

### 2025-05-18 manuscript/manuscript.tex
Add MD repository links

Furthermore, a recent Python version (3.6 or newer) should be available with the following packages installed:
\begin{itemize}
	\item \software{gisttools} (\url{https://github.com/liedllab/gisttools}, $\geq$ 0.2)
	\item \software{mdtraj} (\url{https://www.mdtraj.org/}, $\geq$ 1.9.7) \cite{McGibbon2015-mdtraj}
	\item \software{matplotlib} (\url{https://matplotlib.org/}, $\geq$ 3.7.0) \cite{Hunter2007-matplotlib}
	\item \software{numpy} (\url{https://numpy.org/}, $\geq$ 1.23.5) \cite{harris2020-numpy}
	\item \software{pandas} (\url{https://pandas.pydata.org/}, $\geq$ 1.5.3) \cite{pandas-2023_8239932}
\end{itemize}
If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.
We additionally provide a Jupyter notebook \cite{Kluyver2016-jupyter,Granger2021-jupyter} with the presented analyses. 
For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files, such as \software{PyMol} \cite{pymol} or \software{VMD} \cite{vmd}.
Most of the visualisations and results in this tutorial can also be obtained using the \software{gistpp} program, which is available from \url{https://github.com/KurtzmanLab/Gist-Post-Processing} (code and documentation). \software{Gistpp} can directly manipulate \software{OpenDX} files from the GIST output. 
Its usage and some examples were shown in a previous tutorial on GIST \cite{Ramsey2016}.
For help with the listed cpptraj commands, AMBER atom mask syntax and other AMBER-specific commands, we recommend the following webresources:\\
\begin{itemize}
\item The website https://amberhub.chpc.utah.edu/ which is managed by the Cheatham III group, 
\item The cpptraj manual at https://raw.githubusercontent.com/Amber-MD/cpptraj/master/doc/CpptrajManual.pdf 
\item and the AMBER manual \cite{amber24} at https://ambermd.org/Manuals.php.
\end{itemize}
\pagebreak %TODO: pagebreak, maybe remove these by better formatting

@@

\subsection{Tutorial data}
We will use the crystal structure of streptavidin (PDB code: 1STP \cite{Weber1989-streptavidin-structure}) to start our calculations.
For section \ref{subsec:gist_biotin}, we will use the biotin ligand extracted from the crystal structure.
For the second part in section \ref{subsec:gist_complex}, we will use the whole crystal structure.
For the sake of simplicity, you can download prepared and solvated Amber topologies and structures based on 1STP by running
\begin{lstlisting}[style=bash]
git clone git@github.com:liedllab/gist-tutorial.git
\end{lstlisting}
This will download the tutorial files, including both the manuscript and the examples in the \inlinecode{code} folder.
Additionally, a link to download the original MD trajectory files is provided in the GitHub repository.
\subsection{Running GIST for Biotin}
\label{subsec:gist_biotin}
In this example, we will be running a simple GIST calculation of the biotin solute molecule to get familiar with the usage of GIST in \software{cpptraj}.
\subsubsection{System preparation}
First, extract biotin from the crystal structure and generate a topology and coordinate file compatible with Amber or a simulation package of your choice.
The solute molecule should be solvated with a sufficiently large solvent box. 
For periodic boundary conditions, the size of the solvent box should be large enough for the solvent to behave bulk-like so as to reduce artifacts from solvent-mediated interactions between the periodic images. 
This corresponds to a distance between solute and box boundary of at least \SI{12}{\angstrom} for water \cite{Chen2021,Waibl2022-gist-solvents} though we recommend \SI{15}{\angstrom} or higher.
Carefully check the employed force field, protonation state and partial charges.
%a PDB file containing only biotin. Prepare your structure by adding hydrogens, making sure to choose the correct protonation state. Parametrize biotin with the forcefield of your choice. Solvate the parametrized biotin, making sure to choose a water box of large enough size so that the water on the box edges behaves bulk-like (a wall distance of at least 15\AA is recommended).
A good starting point for the parametrization and solvation of small molecules is the Amber tutorial 4b (\url{http://ambermd.org/tutorials/basic/tutorial4b/}).

>>>>

Furthermore, a recent Python version (3.6 or newer) should be available with the following packages installed:
\begin{itemize}
	\item \software{gisttools} (\url{https://github.com/liedllab/gisttools}, $\geq$ 0.2)
	\item \software{mdtraj} (\url{https://www.mdtraj.org/}, $\geq$ 1.9.7) \cite{McGibbon2015-mdtraj}
	\item \software{matplotlib} (\url{https://matplotlib.org/}, $\geq$ 3.7.0) \cite{Hunter2007-matplotlib}
	\item \software{numpy} (\url{https://numpy.org/}, $\geq$ 1.23.5) \cite{harris2020-numpy}
	\item \software{pandas} (\url{https://pandas.pydata.org/}, $\geq$ 1.5.3) \cite{pandas-2023_8239932}
\end{itemize}
If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.
Additionally, the MD simulation files and raw GIST output files can be found at \url{https://doi.org/10.48323/4mbrd-67m83}, to reproduce the results shown in this tutorial.
We additionally provide a Jupyter notebook \cite{Kluyver2016-jupyter,Granger2021-jupyter} with the presented analyses. 
For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files, such as \software{PyMol} \cite{pymol} or \software{VMD} \cite{vmd}.
Most of the visualisations and results in this tutorial can also be obtained using the \software{gistpp} program, which is available from \url{https://github.com/KurtzmanLab/Gist-Post-Processing} (code and documentation). \software{Gistpp} can directly manipulate \software{OpenDX} files from the GIST output. 
Its usage and some examples were shown in a previous tutorial on GIST \cite{Ramsey2016}.
For help with the listed cpptraj commands, AMBER atom mask syntax and other AMBER-specific commands, we recommend the following webresources:\\
\begin{itemize}
\item The website https://amberhub.chpc.utah.edu/ which is managed by the Cheatham III group, 
\item The cpptraj manual at https://raw.githubusercontent.com/Amber-MD/cpptraj/master/doc/CpptrajManual.pdf 
\item and the AMBER manual \cite{amber24} at https://ambermd.org/Manuals.php.
\end{itemize}
\pagebreak %TODO: pagebreak, maybe remove these by better formatting

@@

\subsection{Tutorial data}
We will use the crystal structure of streptavidin (PDB code: 1STP \cite{Weber1989-streptavidin-structure}) to start our calculations.
For section \ref{subsec:gist_biotin}, we will use the biotin ligand extracted from the crystal structure.
For the second part in section \ref{subsec:gist_complex}, we will use the whole crystal structure.
For the sake of simplicity, you can download prepared and solvated Amber topologies and structures based on 1STP by running
\begin{lstlisting}[style=bash]
git clone git@github.com:liedllab/gist-tutorial.git
\end{lstlisting}
This will download the tutorial files, including both the manuscript and the examples in the \inlinecode{code} folder.
Additionally, the original MD trajectory files and raw GIST output files are provided under the DOI \url{https://doi.org/10.48323/4mbrd-67m83}.
\subsection{Running GIST for Biotin}
\label{subsec:gist_biotin}
In this example, we will be running a simple GIST calculation of the biotin solute molecule to get familiar with the usage of GIST in \software{cpptraj}.
\subsubsection{System preparation}
First, extract biotin from the crystal structure and generate a topology and coordinate file compatible with Amber or a simulation package of your choice.
The solute molecule should be solvated with a sufficiently large solvent box. 
For periodic boundary conditions, the size of the solvent box should be large enough for the solvent to behave bulk-like so as to reduce artifacts from solvent-mediated interactions between the periodic images. 
This corresponds to a distance between solute and box boundary of at least \SI{12}{\angstrom} for water \cite{Chen2021,Waibl2022-gist-solvents} though we recommend \SI{15}{\angstrom} or higher.
Carefully check the employed force field, protonation state and partial charges.
%a PDB file containing only biotin. Prepare your structure by adding hydrogens, making sure to choose the correct protonation state. Parametrize biotin with the forcefield of your choice. Solvate the parametrized biotin, making sure to choose a water box of large enough size so that the water on the box edges behaves bulk-like (a wall distance of at least 15\AA is recommended).
A good starting point for the parametrization and solvation of small molecules is the Amber tutorial 4b (\url{http://ambermd.org/tutorials/basic/tutorial4b/}).
